\documentclass[11pt]{article}
\usepackage[paperwidth=160mm, paperheight=230mm, top=0mm, bottom=15mm, left=20mm, right=20mm]{geometry}
\usepackage{url}
\usepackage{enumitem}
\usepackage{setspace}
\usepackage[hidelinks]{hyperref}

\begin{document}

\pagestyle{empty}

\vspace*{10mm}

\begin{center}
Propositions accompanying the dissertation

\vspace{5mm}

{\Large\bfseries From Insights to Impact: \\ Actionable Interpretability for \\ Neural Machine Translation}

\end{center}

\vspace{5mm}

\begin{enumerate}[leftmargin=*, itemsep=8pt, parsep=0pt]

\item The development of robust and highly usable interpretability tools is a fundamental step towards the development of trustworthy AI systems.

\item Attribution methods can provide a faithful account of contextual influence in the answers of machine translation systems and generative language models.

\item Inference-time interventions on model internals allow for fine-grained and efficient control over the stylistic properties of language model generations.

\item The downstream benefits of machine translation are highly dependent on the resourcedness of source and target languages and their typological similarity.

\item Word-level quality estimation can improve the precision of human post-editing, but render the process more cognitively demanding.

\item The predictive uncertainty of machine translation models can be leveraged to efficiently identify potential errors in their outputs.

\item Human label variation is a critical factor in the evaluation of translation error detection. Annotations from multiple translators lead to more reliable results.

\item Science is a collective truth-seeking effort perpetually in the making.

\item Italians, out of all people, know that \textit{a large field of wheat is slightly pasta}.\footnote{\href{https://gsarti.com/slightly-pasta}{\texttt{gsarti.com/slightly-pasta}}} Our duty is to cook it well, and make sure everyone can enjoy the meal.

\end{enumerate}

\begin{flushright}
Gabriele Sarti \\
December 2025
\end{flushright}

\end{document}